How are galaxies assigned to halos? \\ Searching for assembly bias in the SDSS galaxy clustering

### 2017-03-06 text/ms.tex
apj format

\documentclass[12pt, preprint]{aastex}
%\documentclass[12pt, preprint]{emulateapj}
\usepackage[breaklinks,colorlinks, urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{graphicx}	% For figures
\usepackage{natbib}	% For citep and citep
\usepackage{amsmath}	% for \iint
\usepackage{mathtools}
\usepackage{bm}
\usepackage[breaklinks]{hyperref}	% for blackboard bold numbers
\usepackage{hyperref}
\usepackage{algorithmic,algorithm}
\hypersetup{colorlinks}
\usepackage{color}
\usepackage{morefloats}
\definecolor{dred}{rgb}{0.75,0,0}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

@@

@@

\author{Mohammadjavad Vakili\altaffilmark{1}, Chang~Hoon Hahn\altaffilmark{1}}

\email{mjvakili@nyu.edu}

\altaffiltext{1}{Center for Cosmology and Particle Physics, Department of Physics, New York University, 4 Washington Pl, New York, NY, 10003}

\begin{abstract}

Clustering of dark matter halos has been shown to depend on halo properties beyond mass such as halo concentration, a phenomenon referred to as halo assembly bias. Standard halo occupation modeling (HOD) in large scale structure studies assumes that halo mass alone is sufficient in characterizing the connection between galaxies and halos. Modeling of galaxy clustering can face systematic effects if the number of galaxies are correlated with other halo properties. Using the Small MultiDark-Planck high resolution $N$-body simulation and the measurements of the projected two-point correlation function and the number density of Sloan Digital Sky Survey (SDSS) DR7 main galaxy sample, we investigate the extent to which the dependence of halo occupation on halo concentration can be constrained, and to what extent allowing for this dependence can improve our modeling of galaxy clustering.

@@

\end{abstract}
\keywords{Cosmology: large-scale-structure of the universe, galaxies: halos}

\section{Introduction}

Most theories of cosmology and large-scale structure formation under consideration today rely on the central assumption that galaxies reside in dark matter halos. Detailed study of the galaxy--halo connection is therefore critical in order to constrain cosmological models (by modeling galaxy clustering at non-linear scales) as well as providing a window into galaxy formation physics. One of the most powerful methods for describing the galaxy--halo connection is the halo occupation 
distribution (HOD, see \citealt{seljak2000, berlind_weinberg2002, scoccimarro2001, zheng2005, zheng07, leauthaud12, tinker2013, decorated, 2016arXiv160701782H}).

@@

%In this work, we aim to investigate the dependence of halo occupation on halo concentration, and how this dependence can be constrained in the low-redshift universe with the measurements of the 2PCF of galaxies with a wide range of luminosities in the SDSS DR7 main galaxy sample. 
%%% CHH: Re-phrasing 
In this work, we aim to investigate the dependence of halo occupation on 
{\bf \color{dred} 
    halo concentration and how this dependence can be constrained in the low-redshift universe by 2PCF measurements of galaxies in a wide range of luminosities in the SDSS DR7 main galaxy sample. 
}
%%%
In order to achieve this goal, we need to adopt a HOD model that takes into account 
a dependence on halo properties beyond mass. 
%%% CHH: Re-phrasing to address ref. report
{\bf \color{dred} 
    A number of frameworks in the literature \citep{edHOD-tinker,edHOD-gillmartin,edHOD-weinberg} 
    have proposed environment-dependent HOD models that take into account the large-scale density contrast. 
    In this investigation, we use the following case of the decorated HOD framework \citep{decorated}.  
    In our decorated HOD framework, at fixed halo mass, halos are populated with galaxies according to 
    the standard HOD model. Then using a secondary halo property, halos are split into two populations
    in halo mass bins: halos with the highest and lowest secondary property values. Afterwards, based
    on the assembly bias amplitude parameter, the number of galaxies in the two populations are enhanced
    or reduced. In this model, the assembly bias parameter is not be degenerate with the rest of the HOD 
    parameters. 
}
%%%

@@

For the simulations used in this work, we make use of the Rockstar (\citealt{rockstar}) halo catalogs in the $z=0$ snapshot of the Small MultiDark of Planck cosmology (referred to as $\mathtt{SMDP}$) (\citealt{smallmultidark}). This high resolution $N$-body simulation (publicly available at \url{https://www.cosmosim.org}) was carried out using the 
%%% CHH: small typo
{\bf \color{dred} GADGET-2 (see \citealt{smallmultidark} and references therein) code}, 
%%%
following the Planck $\Lambda$CDM cosmological parameters 
$\Omega_{\rm m} = 0.307$, $\Omega_{\rm b} = 0.048$, $\Omega_{\Lambda} = 0.693$, $\sigma_{8} = 0.823$, $n_{\rm s}=0.96$, 
$h=0.678$. The Box size for this $N$-body simulation is 0.4 $h^{-1} \rm{Gpc}$, the number of simulation particles is 3840$^3$, the mass per simulation particle $m_{\rm p}$ is $9.6 \times 10^{7} \; h^{-1} M_{\odot}$, and the gravitational softening length $\epsilon$ is 1.5 $h^{-1} \rm{kpc}$.

In the $\mathtt{SMDP}$ simulation, as discussed in \citet{halodemographic}, the Rockstar algorithm can reliably resolve halos with $\geq 100$
%%% CHH: grammar error
{\bf \color{dred} particles, which corresponds to $M_{\rm vir} \geq 9.6 \times 10^{9} \; h^{-1}M_{\odot}$.}
%%%
%The advantage of using the $\mathtt{SMDP}$ simulation is that it satisfies both the size and resolution 
%requirements of studying the galaxy--halo connection in a wide range of luminosity thresholds. 
%%%
%%% Re-worded 
{\bf \color{dred}
    The $\mathtt{SMDP}$ simulation provides a number of advantages by satisfying both the size 
    and resolution requirements of studying the galaxy--halo connection in a wide range of 
    luminosity thresholds. For fainter galaxy samples, the faintest galaxies reside in lower 
    mass halos, which requires high resolution. Meanwhile for luminous galaxy samples, their 
    lower number densities requires a large comoving volume. 
}
%%%
Furthermore, since we are studying the higher order halo occupation statistics, the concentration-dependence in particular, it is important to use a simulation that can resolve the internal structure of halos. 
%%% CHH: Re-worded based on (github issue #1)
{\bf \color{dred}
    In the context of Subhalo-Abundance Matching models, which requires subhalo completeness 
    in the low mass limit, the $\mathtt{SMDP}$ simulation has been used to model the faintest 
    galaxy samples in the SDSS data~\citep[see][]{hod_vs_sham}. 
}
%%%
The added advantage of using the $\mathtt{SMDP}$ simulation over some of the other industrial simulation boxes commonly used in the literature such as $\mathtt{Bolshoi}$ (\citealt{Klypin2011,smallmultidark}) simulation is its larger comoving volume. Larger volume makes this simulation more suitable for performing inference with $L_{\star}$ (corresponding to $M_{\rm r} \sim -20.44$, see \citealt{blanton2003}) and more luminous than $L_{\star}$ galaxy samples that occupy larger comoving volumes.

@@

For populating the halos with galaxies, we follow the procedure described in \citet{2016arXiv160701782H}, and \citet{decorated}. The central galaxies are assumed to be at the center of the host dark matter halos. We assume that the central galaxies are at rest with respect to the bulk motion of the halos and their velocities are given by the velocity of the center of mass of their host halo. Note that this assumption is shown to be violated in brighter than $L_{\star}$ galaxy samples (see \citealt{guo2015}). But since we are not considering the redshift space 2PCF multipoles in our study, we do not expect this velocity bias to impact our inference. We place the satellite galaxies within the virial radius of the halo following a Navarro-Frenk-White profile (hereafter NFW; \citealt{nfw}). This approach is different from other simulation-based halo occupation modeling techniques (see \citealt{hod_vs_sham,zheng_guo}) in that the positions of the satellites are not assigned to the dark matter particles in the $N$-body simulation. 
%%% Missing citation
{\bf \color{darkgreen} 
    The concentration of the NFW profile is given by the empirical 
    mass-concentration relation provided by \citet{nfw_c(M)}. 
}
%%%
The velocities of the satellite galaxies are given by two components. The first component is the velocity of the host halo. The second component is the velocity of the satellite galaxy with respect to the host halo which is computed following the solution to the NFW profile Jeans equations (\citealt{more2010}). We refer the readers to \citet{decorated} for a more comprehensive and detailed discussion of the forward modeling of the galaxy mock catalogs.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WP MODEL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[p]~\\
\begin{center}
\includegraphics[width=\textwidth]{wpmodel.pdf}
\caption{Comparison between the posterior predictions of $w_{p}(r_{p})$ and the SDSS $w_{p}(r_{p}$ measurements. Predictions from the standard HOD model (HOD model with assembly bias) are shown in red (blue). The Dark and light shaded regions mark the 68$\%$ and the 95$\%$ confidence intervals. The errorbars are from the diagonal elements of the covariance matrix.}
\label{fig:wpmodel}
\end{center}
\end{figure*}
\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WPRES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[p]~\\
\begin{center}
\includegraphics[width=\textwidth]{wpres.pdf}
\caption{Same as Figure \ref{fig:wpmodel}, but showing the fractional difference between the posterior predictions and the observed projected 2PCF for all the luminosity threshold samples. In all luminosity threshold samples, predictions of the two models for small scale clustering are consistent. In the samples that favor more positive values of the central assembly bias parameter ($M_{\rm r}<-19.5,-19,-20,-20.5$), modeling of the intermediate and large scale clustering is slightly improved. The large scale clustering modeling of the $M_{\rm r}<-18$ sample is also improved because of negative constraints on $\acen$ which is equivalent to allocation of more central galaxies in low concentration halos at fixed halo mass.}
\label{fig:wpres}
\end{center}
\end{figure*}

@@

@@

%%
%% End of file `sample.tex'.

>>>>

\documentclass[twocolumn]{aastex61}
\pdfoutput=1 %for arXiv submission
\usepackage{amsmath,amstext}
\usepackage[T1]{fontenc}
\usepackage{apjfonts} 
\usepackage[figure,figure*]{hypcap}

\renewcommand*{\sectionautorefname}{Section} %for \autoref
\renewcommand*{\subsectionautorefname}{Section} %for \autoref

\definecolor{dred}{rgb}{0.75,0,0}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{ colorlinks,
linkcolor=darkblue,
filecolor=darkgreen,
urlcolor=darkred,
citecolor=darkblue}

@@

\shorttitle{AASTeX 6.1 Template}
\shortauthors{Vakili \& Hahn}

@@

\author{Mohammadjavad Vakili}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 4 Washington Pl, New York, NY, 10003}

\author{ChangHoon Hahn}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 4 Washington Pl, New York, NY, 10003}

\begin{abstract}
Clustering of dark matter halos has been shown to depend on halo properties beyond mass such as halo concentration, a phenomenon referred to as halo assembly bias. Standard halo occupation modeling (HOD) in large scale structure studies assumes that halo mass alone is sufficient in characterizing the connection between galaxies and halos. Modeling of galaxy clustering can face systematic effects if the number of galaxies are correlated with other halo properties. Using the Small MultiDark-Planck high resolution $N$-body simulation and the measurements of the projected two-point correlation function and the number density of Sloan Digital Sky Survey (SDSS) DR7 main galaxy sample, we investigate the extent to which the dependence of halo occupation on halo concentration can be constrained, and to what extent allowing for this dependence can improve our modeling of galaxy clustering.

@@

\end{abstract}

\keywords{Cosmology: large-scale-structure of the universe--- galaxies: halos}

\section{Introduction}
Most theories of cosmology and large-scale structure formation under consideration today rely on the central assumption that galaxies reside in dark matter halos. Detailed study of the galaxy--halo connection is therefore critical in order to constrain cosmological models (by modeling galaxy clustering at non-linear scales) as well as providing a window into galaxy formation physics. One of the most powerful methods for describing the galaxy--halo connection is the halo occupation 
distribution (HOD, see \citealt{seljak2000, berlind_weinberg2002, scoccimarro2001, zheng2005, zheng07, leauthaud12, tinker2013, decorated, 2016arXiv160701782H}).

@@

%In this work, we aim to investigate the dependence of halo occupation on halo concentration, and how this dependence can be constrained in the low-redshift universe with the measurements of the 2PCF of galaxies with a wide range of luminosities in the SDSS DR7 main galaxy sample. 
%%% CHH: Re-phrasing 
In this work, we aim to investigate the dependence of halo occupation on 
{\bf \color{dred} 
    halo concentration and how this dependence can be constrained in the low-redshift universe by 2PCF measurements of galaxies in a wide range of luminosities in the SDSS DR7 main galaxy sample. 
}
%%%
In order to achieve this goal, we need to adopt a HOD model that takes into account 
a dependence on halo properties beyond mass. 
%%% CHH: Re-phrasing to address ref. report
{\bf \color{red} 
    A number of frameworks in the literature \citep{edHOD-tinker,edHOD-gillmartin,edHOD-weinberg} 
    have proposed environment-dependent HOD models that take into account the large-scale density contrast. 
    In this investigation, we use the following case of the decorated HOD framework \citep{decorated}.  
    In our decorated HOD framework, at fixed halo mass, halos are populated with galaxies according to 
    the standard HOD model. Then using a secondary halo property, halos are split into two populations
    in halo mass bins: halos with the highest and lowest secondary property values. Afterwards, based
    on the assembly bias amplitude parameter, the number of galaxies in the two populations are enhanced
    or reduced. In this model, the assembly bias parameter is not be degenerate with the rest of the HOD 
    parameters. 
}
%%%

@@

For the simulations used in this work, we make use of the Rockstar (\citealt{rockstar}) halo catalogs in the $z=0$ snapshot of the Small MultiDark of Planck cosmology (referred to as $\mathtt{SMDP}$) (\citealt{smallmultidark}). This high resolution $N$-body simulation (publicly available at \url{https://www.cosmosim.org}) was carried out using the 
%%% CHH: small typo
{\bf \color{red} GADGET-2 (see \citealt{smallmultidark} and references therein) code}, 
%%%
following the Planck $\Lambda$CDM cosmological parameters 
$\Omega_{\rm m} = 0.307$, $\Omega_{\rm b} = 0.048$, $\Omega_{\Lambda} = 0.693$, $\sigma_{8} = 0.823$, $n_{\rm s}=0.96$, 
$h=0.678$. The Box size for this $N$-body simulation is 0.4 $h^{-1} \rm{Gpc}$, the number of simulation particles is 3840$^3$, the mass per simulation particle $m_{\rm p}$ is $9.6 \times 10^{7} \; h^{-1} M_{\odot}$, and the gravitational softening length $\epsilon$ is 1.5 $h^{-1} \rm{kpc}$.

In the $\mathtt{SMDP}$ simulation, as discussed in \citet{halodemographic}, the Rockstar algorithm can reliably resolve halos with $\geq 100$
%%% CHH: grammar error
{\bf \color{red} particles, which corresponds to $M_{\rm vir} \geq 9.6 \times 10^{9} \; h^{-1}M_{\odot}$.}
%%%
%The advantage of using the $\mathtt{SMDP}$ simulation is that it satisfies both the size and resolution 
%requirements of studying the galaxy--halo connection in a wide range of luminosity thresholds. 
%%%
%%% Re-worded 
{\bf \color{red}
    The $\mathtt{SMDP}$ simulation provides a number of advantages by satisfying both the size 
    and resolution requirements of studying the galaxy--halo connection in a wide range of 
    luminosity thresholds. For fainter galaxy samples, the faintest galaxies reside in lower 
    mass halos, which requires high resolution. Meanwhile for luminous galaxy samples, their 
    lower number densities requires a large comoving volume. 
}
%%%
Furthermore, since we are studying the higher order halo occupation statistics, the concentration-dependence in particular, it is important to use a simulation that can resolve the internal structure of halos. 
%%% CHH: Re-worded based on (github issue #1)
{\bf \color{red}
    In the context of Subhalo-Abundance Matching models, which requires subhalo completeness 
    in the low mass limit, the $\mathtt{SMDP}$ simulation has been used to model the faintest 
    galaxy samples in the SDSS data~\citep[see][]{hod_vs_sham}. 
}
%%%
The added advantage of using the $\mathtt{SMDP}$ simulation over some of the other industrial simulation boxes commonly used in the literature such as $\mathtt{Bolshoi}$ (\citealt{Klypin2011,smallmultidark}) simulation is its larger comoving volume. Larger volume makes this simulation more suitable for performing inference with $L_{\star}$ (corresponding to $M_{\rm r} \sim -20.44$, see \citealt{blanton2003}) and more luminous than $L_{\star}$ galaxy samples that occupy larger comoving volumes.

@@

For populating the halos with galaxies, we follow the procedure described in \citet{2016arXiv160701782H}, and \citet{decorated}. The central galaxies are assumed to be at the center of the host dark matter halos. We assume that the central galaxies are at rest with respect to the bulk motion of the halos and their velocities are given by the velocity of the center of mass of their host halo. Note that this assumption is shown to be violated in brighter than $L_{\star}$ galaxy samples (see \citealt{guo2015}). But since we are not considering the redshift space 2PCF multipoles in our study, we do not expect this velocity bias to impact our inference. We place the satellite galaxies within the virial radius of the halo following a Navarro-Frenk-White profile (hereafter NFW; \citealt{nfw}). This approach is different from other simulation-based halo occupation modeling techniques (see \citealt{hod_vs_sham,zheng_guo}) in that the positions of the satellites are not assigned to the dark matter particles in the $N$-body simulation. 
%%% Missing citation
{\bf \color{green} 
    The concentration of the NFW profile is given by the empirical 
    mass-concentration relation provided by \citet{nfw_c(M)}. 
}
%%%
The velocities of the satellite galaxies are given by two components. The first component is the velocity of the host halo. The second component is the velocity of the satellite galaxy with respect to the host halo which is computed following the solution to the NFW profile Jeans equations (\citealt{more2010}). We refer the readers to \citet{decorated} for a more comprehensive and detailed discussion of the forward modeling of the galaxy mock catalogs.

@@

%\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WP MODEL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[p]~\\
\begin{center}
\includegraphics[width=\textwidth]{wpmodel.pdf}
\caption{Comparison between the posterior predictions of $w_{p}(r_{p})$ and the SDSS $w_{p}(r_{p}$ measurements. Predictions from the standard HOD model (HOD model with assembly bias) are shown in red (blue). The Dark and light shaded regions mark the 68$\%$ and the 95$\%$ confidence intervals. The errorbars are from the diagonal elements of the covariance matrix.}
\label{fig:wpmodel}
\end{center}
\end{figure*}
%\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WPRES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[p]~\\
\begin{center}
\includegraphics[width=\textwidth]{wpres.pdf}
\caption{Same as Figure \ref{fig:wpmodel}, but showing the fractional difference between the posterior predictions and the observed projected 2PCF for all the luminosity threshold samples. In all luminosity threshold samples, predictions of the two models for small scale clustering are consistent. In the samples that favor more positive values of the central assembly bias parameter ($M_{\rm r}<-19.5,-19,-20,-20.5$), modeling of the intermediate and large scale clustering is slightly improved. The large scale clustering modeling of the $M_{\rm r}<-18$ sample is also improved because of negative constraints on $\acen$ which is equivalent to allocation of more central galaxies in low concentration halos at fixed halo mass.}
\label{fig:wpres}
\end{center}
\end{figure*}

%\clearpage

@@

%\clearpage

@@
